An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)

### 2022-04-19 manuscript/main.tex
updated manuscript

Matrix reasoning tasks are among the most well-studied and useful measures in the behavioral sciences. Matrix reasoning requires participants to use abstract reasoning to solve puzzles. Matrix reasoning indexes fluid intelligence and is correlated with working memory capacity \citep{conway2002latent, kane2004generality, unsworth2005working, salthouse2014relations}. In low-stakes testing environments, performance on the progressive matrices is associated with participant motivation and willingness to expend effort \citep{gignac2018moderate, gignac2019maximum}. The progressive matrices has excellent external validity, and is predictive of academic performance \citep{roth2015intelligence} and scores on college admission exams \citep{frey2004scholastic, koenig2008act}. % Thus, matrix reasoning tasks are increasingly utilized as control measure in individual difference behavioral studies \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

Unfortunately there are a number of challenges in using matrix reasoning tasks as part of behavioral research. One challenge is the problem of copyright. The most prominent matrix reasoning tasks, such as the Raven's progressive matrices \citep{raven2003raven} and the WAIS matrix subtest \citep{wechsler1999wechsler}, are typically not free to use and have restrictions against being computerized. A second challenge is that, across the measures in the public domain, there are relatively few items available. For example, the Hagen Matrices Test \citep{heydasch2014hagen} and ICAR matrix reasoning test \citep{condon2014international} have only 20 and 16 items, respectively. The availability of only a limited number of items raises the possibility of practice effects \citep{ng1974applicability, bors2003effect}, especially in the age of online experiments where multiple groups of researchers may be unwittingly administering the same test to many of the same participants.

In response to these challenges, \cite{chierchia2019matrix} developed, and released into the public domain, the matrix reasoning item bank (MaRs-IB). The MaRs-IB includes 80 matrix reasoning puzzles. Each puzzle consists of a 3x3 matrix containing abstract shapes in eight out of nine cells. To solve a puzzle, participants must deduce which of four response options completes the matrix based on the pattern of changing relations among the shapes across cells (Figure 1). The puzzles in the MaRs-IB vary considerably in their complexity, both in the number of shapes per cell and in the number of changing relations across shapes. Importantly each template puzzle has six clones, which are matched in terms of complexity but differ in their perceptual properties (one of three shape sets) and their response options (one of two distractor sets). Thus, the MaRs-IB offer considerable utility insofar that they are free to use, have many items, and many unique items to prevent practice effects.

A preliminary investigation of the psychometric properties of the MaRs-IB items proved promising. Indeed, \cite{chierchia2019matrix} found that scores on a short-form matrix reasoning test using MaRs-IB items had good split-half reliability and test-retest reliability. The authors also found that performance observed scores on this test were moderately correlated with a digit span task and the ICAR matrix reasoning test, indicating good convergent validity. % note the release of summary statistics here?

Unfortunately, there are two notable limitations in this initial investigation. The first is that, due to the design of the matrix reasoning test, there is little data available for the majority of the items in the MaRs-IB. Just over half the items (42/80) were completed by 100 or more participants. As such, there is too little data to characterize the psychometric properties of each item with reasonable confidence (an issue that is exacerbated when considering each template item has six variants). A second issue is that, also due to the design of the matrix reasoning test, participants were allowed to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, items later in the sequence are likely to be contaminated by participants sacrificing accuracy for speed, thereby distorting the summary statistics for these items.\footnote{appendix: Indeed, looking only at the easiest MaRs-IB items (dimension = 1), there is a strong positive correlation between number of completing participants and mean accuracy ($r = 0.673$).} Thus, caution is warranted in interpreting the results of the original study and additional validation is warranted.

%Better intro sentence tk: mention that this is a more extensive psychometric validation of the MaRs-IB in a much larger dataset, with three major improvements.
The most important contribution of the current work is the evaluation of the psychometric properties of (almost) every item in the MaRs-IB using item response theory (IRT) \citep{embretson2013item, de2013theory}. IRT models confer several important benefits for the purposes of psychometric validation. First they provide an estimate of the difficulty and discrimination of each item, where the latter is an index of how informative or reliable an item is. Through these parameters, IRT models allows for the computation of the test information function (TIF), which measures how sensitive a test is at measuring performance at varying levels of ability. Finally, IRT make possible optimal test assembly \citep{van1998optimal}, or design of new tests with maximal sensitivity to measuring performance in a specified range of abilities under varying constraints (e.g. test length, time limits, parallel forms with equal TIF).

A second contribution offered here is an analysis of the relationships between attributes of MaRs-IB items and their psychometric properties using explanatory item response models \citep{de2004explanatory, wilson2008explanatory}. Explanatory IRT models provide a framework for measuring how item parameters (e.g. difficulty) are shaped by their features. As such, these models provide means to evaluate how successful was the generation of the items in the MaRs-IB. For example, we can quantify the extent to which features of item complexity (e.g. number of elements, number of changing relations) predict item difficulty. Moreover, we can determine whether two items that vary only in their distractor set are psychometrically equivalent. Explanatory IRT models have the additional potential benefit of producing more accurate estimates of item parameters \citep{neuhaus2006separating}.

%item clones would be isomorphic (psychometrically equivalent). In any realistic situation, the 2P-F is misspecified as it ignores the psychometric deviations introduced in the item cloning process (Glas & van der Linden, 2003).

A third contribution is the decomposition of variance in item functioning from modeled and unmodeled sources using additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. This framework provides another means by which to evaluate the success of the design of the MaRs-IB by quantifying the residual variance in item parameters after accounting for item attributes. For example, two items of equal complexity may turn out not be psychometrically equivalent if unmodeled attributes (e.g. the types of changing relations) implicitly impact item functioning. Similarly, two item clones may not be psychometrically equivalent due to differences in their superficial properties. Quantifying the degree of error in the item cloning process is critical, as the degree of nonequivalent dictates whether clones are exchangeable. That is, if clones are not psychometrically equivalent, then they cannot be used interchangeably in parallel short forms.

The paper proceeds as follows. First we analyze participants' response data hierarchical item response models, which allowed us to measure all items' difficulty and discrimination while controlling for participants' varying abilities. The item response models also allowed us to measure directly the extent to which that item variants are truly equivalent. We then demonstrate how these item response parameters can be used to construct new short-forms with desirable properties. Using these methods, we introduce three new MaRs-IB short-form tests with good internal consistency and convergent validity.

All data and code, including items' estimated parameters and example code for test assembly, is available at: \url{https://github.com/szorowi1/mars-irt}.

\section{Calibration Study}

@@

The purpose of the first study was to calibrate the items of the MaRs-IB, using response data collected from a large number of adult participants from the general population. In particular, we sought to accomplish the following three objectives: (1) to characterize the psychometric properties of the items in the MaRs-IB using item response models; (2) to investigate the associations between item attributes and their psychometric properties, and (3) to determine the equivalence of items with similar complexities, and clones of the same item templates.

@@

To ensure data quality, the data from multiple participants were excluded prior to analysis (see Exclusion Criteria below) leaving the data from N=1501 participants for analysis. The majority of participants in this sample identified as women (men: N=670; women: N=811; non-binary or other: N=13; rather not say: N=2) and were 28.7 years old (sd = 9.9) on average. The sample was relatively well-educated, with the majority having completed a Bachelor's degree (N=507) or Master's degree or higher (N=322). By comparison, fewer participants endorsed having completed only some college (N=471), only a high school degree (N=199), or preferred not to say (N=2).

@@

After providing consent, participants completed 16 items from the MaRs-IB. The design of the MaRs-IB items have been described previously \citep{chierchia2019matrix}. Briefly, each MaRs-IB item consists of a 3 x 3 matrix. Eight of the nine resulting cells contain abstract shapes, while one cell on the bottom right-hand side of the matrix is empty. Participants are instructed to ``complete the matrix'' by finding the missing shape from among four possible alternatives.

Each trial of the experiment began with a fixation cross, lasting  1200 ms, after which a MaRs-IB item was presented. Participants were then given up to 30 s to solve the puzzle. After 25 s a clock appeared, indicating that 5 s remained before presentation of the next item. A trial ended when participants responded, or after 30 s had elapsed without response. Before the trials began, participants reviewed instructions and were made to correctly complete three practice items. Participants were told to respond carefully but to guess before the timer ran out. The format of the experiment and instructions were adapted from code made publicly available by the original authors (\url{https://app.gorilla.sc/openmaterials/36164}).

In order to ensure sufficient and uniform sampling of the MaRs-IB items, all participants completed 16 items pseudorandomly selected from the 64 available MaRs-IB items.\footnote{We did not collect new data for the 16 easiest items from the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) as the original study or pilot testing found performance on these items to be at or near ceiling.} The item sets were constructed as follows: First, we subdivided the 64 items into 16 sets of four based on their complexity. Participants were then randomly assigned one item from each of the 16 sets. In this way, all participants received test forms of approximately equal difficulty. We also counterbalanced which variant of each item was presented across participants, such that we had approximately the same number of responses available by shape set (1, 2, or 3) and distractor type (minimal difference or or paired difference; see \cite{chierchia2019matrix} for details).

@@

The data from multiple participants who completed the experiment were excluded prior to analysis for one or more of the following reasons: rapid guessing (responding in less than 3 s) on four or more trials (N=70); failing to respond on four or more trials (N=8); or minimizing the browser window to a size smaller than the minimally required dimensions (N=7). In total 83 of 1584 (5.2\%) participants were excluded, leaving the data from N=1501 participants available for analysis. Among these participants, each item template (64 in total) was sampled approximately 375 times (sd = 12) and each item clone (384 in total) was sampled approximately 62 times (sd = 7).

@@

We employed item response models to characterize the psychometric properties of each item. Specifically, we used additive multilevel item structure (AMIS) models with random residuals \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. In item structure models, item clones are modeled as descending from a template item. The basis of all of these models was the three-parameter logistic (3PL) item response model, where the probability of a correct response ($y_{ijk} = 1$) for person $i$ on item clone $k$ belonging to item family $j$ is:

@@

\noindent where $\theta_i$ is the latent ability for person $i$, and $\beta_{jk}$, $\alpha_{jk}$, and $\gamma_{jk}$ are the difficulty, discrimination, and guessing parameters for item clone $k$ of item family $j$. Because guessing parameters are challenging to reliably estimate in the absence of very large amounts of data \citep{han2012fixing}, we fixed the guessing parameter for each item clone to the nominal guessing rate ($\gamma_{jk} = 0.25$).

@@

\noindent To elaborate, each item difficulty parameter was modeled as an intercept ($\mu_\beta$), reflecting the average difficulty across all items, and four additional components:

@@

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of the template (level-1) attributes on item difficulty, where $Q_{jn}$ is the value of attribute $n$ for item template $j$, and $\delta_{\beta n}$ is the effect of attribute $n$.

\item $\epsilon_{\beta j}$: the template (level-1) residual, or residual variability in item difficulty of the item template unexplained by its attributes.

\item $\sum_{m=1}^M R_{km} \delta_{\beta m}$: the effect of the clone (level-2) attributes on item difficulty, where $R_{km}$ is the value of attribute $m$ for item clone $k$, and $\delta_{\beta m}$ is the effect of attribute $m$.

\item $\epsilon_{\beta k}$: the clone (level-2) residual, or residual variability in item difficulty of the item clone unexplained by its attributes.

@@

Similarly, item discrimination parameters were also modeled as the sum of an equivalent set of components:

@@

The AMIS model framework provides a straightforward means of achieving Aims 2 and 3. Via the attribute effect components, we can test for any associations between item attributes and parameters (Aim 2). So too, via the error terms, we can quantify the magnitude of residual variance in item parameters from unmodeled sources both at the level of item templates and item clones (Aim 3). If the residual variance in item parameters across item clones is sufficiently large, then we can conclude item clones are neither psychometrically equivalent nor exchangeable.

In the following analyses, we consider two template (level-1) attributes: feature number, or the number of unique shapes in each cell of an item; and rule number, or the number of changing relationships among the shapes (e.g. color change, shape change, position change) a participant must consider in order to solve the item. We predicted that both attributes would be positively related to item difficulty. In contrast, we had no hypotheses about the relationship between these attributes and item discrimination.

In turn, we consider two clone (level-2) attributes: distractor type, or MD-type distractors versus PD-type distractors; and mean response time, or the average time spent deliberating on that item. Based on the preliminary findings of \cite{chierchia2019matrix}, we predicted there would be no association between distractor type and either type of item parameter. We note that, in contrast to all of the other modeled attributes, average response time is not an intrinsic property of an item. We opted to include it in our models, however, because of its previously established relationship with item difficulty \citep{chierchia2019matrix}, and because including response time as a covariate has previously been shown to improve parameter estimation \citep{bertling2018using}. To prevent any collinearity between the attributes, mean response time was orthogonalized with respect to the three other item attributes, and therefore reflects any residual structure after accounting for all other modeled effects.\footnote{Note that we intentionally do not include shape set, or the superficial properties of an item, as covariates in our model. This is due to the would-be complexity of doing so. Though each item has three clones, the clones do not all have the same perceptual patterns. There are XX unique patterns. Future studies might consider even larger datasets to study perceptual influence.}

Finally, we estimated person-level abilities following an explanatory item response modeling approach \citep{wilson2008explanatory}:

@@

@@

During estimation, the item discrimination parameters were restricted to be in the range $\alpha_{jk} \in [0, 5]$.

To ensure the identifiability of all models, person abilities were constrained to have a mean of zero and a variance of 1.0, such that the residual variance was defined as $V(\epsilon) = 1 - \sum \rho_p^2$. Note that all person-level attributes were standard-scored prior being entered into the model.

@@

The goodness of fit of the models was compared using Bayesian leave-one out cross-validation \citep{vehtari2017practical}, which has been found to perform better than more traditional information criteria for comparing item response models \citep{luo2017performances}. Importantly, we computed the conditional leave-one-cluster out (LOCO) cross- validation \citep{merkle2019bayesian}, which measures a model's ability to generalize to held out items (rather than held out responses or held out participants) --- i.e., for inferring generalization to additional items constructed using the same features rather than to additional participants sampled from the same population.

@@

The fit of the best-fitting model to the data was evaluated through posterior predictive model checks \citep{gelman1996posterior, levy2017bayesian}. Item fit was assessed using the Orlando and Thissen index ($S-\chi^2$) \citep{toribio2011discrepancy}, which compares the observed and expected proportions of participants who answer an item correctly at each raw-score group. The unidimensionality of the data was assessed using the generalized dimensionality discrepancy measure (GDDM) \citep{levy2011generalized}, which tests for violation of the assumption of local independence. This latter measure is crucial for ensuring there are negligible levels of local item dependence in the data, which may otherwise lower the efficiency of future test forms due to the inclusion of correlated, redundant items.

@@

On average, participants completed 9.6 of 16 items correctly (sd = 3.1, IQR = 8.0 - 12.0) (Figure \ref{fig:fig01}a). Timing out occurred on only 1.8\% of trials. A total of 43 participants (2.9\%) performed below chance (i.e. fewer than four items correct), while only 11 participants (0.7\%) correctly responded to all 16 items. Thus, over 95\% of participants' observed scores were within a reasonable range. These results corroborate \cite{chierchia2019matrix} in that there were no obvious ceiling or floor effects in performance, and that the majority of participants were sufficiently motivated to participate despite the low-stakes testing environment.

Across all 384 unique item variants, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74) (Figure \ref{fig:fig01}b). A total of 12 items (3.1\%) had accuracy rates beneath chance levels, and only 22 items (5.7\%) had performance levels above 90\%. Mean performance on an item was significantly and negatively correlated with the number of changing relationships amongst its features (Spearman's $\rho = -0.482, p < 1e^{-6})$). Moreover, item-wise accuracy rates in this sample were remarkably similar to those reported by \cite{chierchia2019matrix} (Spearman's $\rho = 0.870, p < 1e^{-6})$), when restricting the latter to items with more than 200 available responses. These results further corroborate \cite{chierchia2019matrix} in that the majority of items exhibited good performance, with few items at ceiling or floor and passing of the manipulation check.

We investigated the relationship between response accuracy and times using a random-intercepts linear regression model. Participants' log-transformed, item-level response times were predicted as a function of item accuracy, rest score (observed score for all other items), and item difficulty (one minus the proportion correct on that item). The results are summarized in Table XX. On average, participants spent 15.9 s (sd = 7.2 s) on each item. There were significant main effects of accuracy ($\beta = 0.067$, $t = 9.84$, $p < 0.001$), rest score ($\beta = 0.125$, $t = 16.95$, $p < 0.001$), and item difficulty ($\beta = 0.137, t = 44.33, p < 0.001$) such that response times were slower for correct responses, better-performing participants, and more difficult items.

There was also a significant accuracy by difficulty interaction (correct responses were even slower for more difficult items), and a significant accuracy by score interaction (correct responses were faster for better performing participants). Finally, there was a significant difficulty by score interaction, such that better performing participants exhibited slower responses for more difficult items. This pattern of response time results is visible in Figure \ref{fig:fig01}c. The three-way interaction term was not statistically significant.

@@

\item Model 1a: Item difficulty is a function only of template attributes (number of shapes, number of changing relations). In this model, items of equal complexity are psychometrically equivalent as are all item clones.

>>>>

Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences. Much of their popularity undoubtedly reflects their versatility in use. Matrix reasoning tasks are strong (albeit impure) indicators of general intelligence \citep{gignac2015raven} and working memory capacity \citep{kane2004generality, unsworth2005working}. They are predictive of important real-world outcomes such as childhood academic performance \citep{roth2015intelligence} and performance on college entrance exams \citep{frey2004scholastic, koenig2008act}. In low-stakes testing settings, matrix reasoning tasks can additionally function as measures of motivation, willingness to expend mental effort, and other aspects of personality \citep{duckworth2011role, gignac2019maximum}. Performance on matrix reasoning tasks can also serve as an instrumental variable in studies of psychiatric populations; that is, it can be used to control for general disruptions to cognitive ability where specific domains of cognition are primarily of interest \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

Unfortunately there are a number of impediments to using matrix reasoning tasks as part of behavioral research. One such challenge is the problem of copyright. The most prominent examples of matrix reasoning tasks, such as Raven's progressive matrices \citep{raven2003raven} and the WAIS matrix subtest \citep{wechsler1999wechsler}, are not free to use and have legal restrictions against being computerized. A second challenge is that, across all of the matrix reasoning tests in the public domain, there are relatively few unique items available. The Hagen matrices test \citep{heydasch2014hagen} and ICAR matrix reasoning test \citep{condon2014international}, for example, have only 20 and 16 items respectively. The availability of only a limited number of items raises the possibility of repeat exposure effects, which threaten the validity of these measures  \citep{ng1974applicability, bors2003effect}. This problem is exacerbated in the current age of online experiments, where multiple groups of researchers may be unintentionally administering the same test to the same participants recruited from the same online labor platforms.

Noting these challenges, \cite{chierchia2019matrix} developed and made publicly available the matrix reasoning item bank (MaRs-IB), a collection of 80 unique matrix reasoning puzzles. Each puzzle in the MaRs-IB consists of a 3x3 matrix containing abstract shapes in eight out of nine cells (Figure \ref{fig:fig00}). To solve a puzzle, participants must deduce which of four response options correctly completes the matrix based on the pattern of changing relations among the shapes across cells. The puzzles in the MaRs-IB were designed to span a considerable range of complexity, both in the number of unique shapes and changing relations in and across cells. To further increase the reusability of the MaRs-IB and minimize practice effects, each puzzle (hereafter referred to as a template item) has six clones. An item's clones share identical structural properties (i.e. number of shapes, number of changing relations) but may vary in their distractors (two unique sets per item) or perceptual features (three unique shape sets per item). Thus, the MaRs-IB addresses several issues associated with other matrix reasoning tests.

\cite{chierchia2019matrix} also conducted an initial study of the psychometric properties of the MaRs-IB. A sample of 659 participants completed a fixed-length and fixed-order test measure in which they had eight minutes to complete as many MaRs-IB items as possible. The authors found that performance on this test measure had good split-half and test-retest reliability. Moreover they found that participants' ability to solve puzzles in the MaRs-IB was moderately predictive of their performance on a digit span task and the ICAR matrix reasoning test, indicating good convergent validity. At the conclusion of this study, \cite{chierchia2019matrix} publicly released a summary of the difficulty (i.e. proportion correct responses) of each item in the MaRs-IB so that future researchers could construct new test measures of custom difficulty and duration.

Unfortunately, there are two critical limitations in this study that undermine the utility of the summary statistics. The first is that, given design of the test, there was relatively little data collected for the majority of the items in the MaRs-IB. Indeed, only half (42 of 80 items) were completed by 100 or more participants. As such, one cannot be confident about the reported functioning of a majority of the items (an issue that is exacerbated when considering that each item has six clones). A second issue is participants were allowed to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, the response data for items that came later in the fixed-order test were increasingly likely to be produced by participants sacrificing accuracy for speed, thereby biasing the proportion correct measure for these items (see the Supplementary materials for evidence of this point). Therefore, additional investigation of the functioning and psychometric properties of the MaRs-IB is warranted.

Here we provide a more extensive study and psychometric validation of the MaRs-IB. The most important contribution of the current work is the characterization of (almost) every item in the MaRs-IB using item response theory (IRT) \citep{embretson2013item, de2013theory}. IRT models confer several important advantages for the purposes of psychometric validation. First they provide an estimate of the difficulty and discrimination of each item, where the latter is an index of how much information an item provides given a level of ability. Using these quantities, IRT models in turn facilitate the computation of the test information function (TIF), which measures the reliability of a test measure across a range of ability. Finally, IRT models make possible optimal test assembly  \citep{van1998optimal}, or the design of new tests with maximal reliability given researcher-specified constraints (e.g. test duration, test difficulty).

A second contribution provided here is an investigation of the relationship between the attributes of MaRs-IB items and their psychometric properties using explanatory item response models \citep{de2004explanatory, wilson2008explanatory}. The utility of the MaRs-IB depends in part on its items exhibiting certain expected characteristics. Across items we should expect complexity (e.g. more shapes per cell, more changing relations) to be positively associated with difficulty. Moreover if item clones are exchangeable (i.e. perfect replicas with identical psychometric properties), then we should expect no association between distractor type and difficulty. Thus explanatory IRT models offer an opportunity to evaluate the success of the MaRs-IB design. As a secondary benefit, explanatory IRT models also provide more accurate estimates of item parameters \citep{neuhaus2006separating}.

A third contribution is the decomposition of variance in item functioning, into modeled and unmodeled sources, using additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. First we use AMIS models to quantify the residual variance in difficulty and discrimination across items after accounting for their respective attributes. At the item-level, smaller residual variances indicate that items are functioning as intended. We also use AMIS models to quantify the variance in difficulty and discrimination across item clones. If item clones are to be treated as exchangeable (i.e. able to be used interchangeably), then the residual variance in their item parameters must be negligible. AMIS models therefore provide a second means by which to evaluate the success of the MaRs-IB design.

The remainder of the paper proceeds as follows. First we report a calibration study, in which a large sample of participants each completed a subset of items from the MaRs-IB. Using their response data, we then fit a series of multilevel item structure models in order to estimate item parameters for each item (and their clones). Using these same models, we also interrogate the relationship between item attributes and functioning within and cross items in the MaRs-IB. Then we report a second validation study, in which a smaller sample of participants completed one of three new MaRs-IB short-form tests. These tests --- designed using optimal test assembly methods applied to the item parameters estimated during the calibration study --- were found to have good psychometric properties and convergent validity. This second study provides a blueprint for how to construct new MaRs-IB test measures using the item parameter estimates that have been released as part of this study.

All data, code, and model outputs (including the item parameter estimtes) are publicly available at: \url{https://github.com/dawlab/mars-irt}.

\clearpage
\section{Calibration Study}

@@

The purpose of the first study was to calibrate the items of the MaRs-IB using response data collected from a large number of adult participants from the general population. In particular, we sought to accomplish the following three objectives: (1) to characterize the psychometric properties of the items in the MaRs-IB using item response models; (2) to investigate the associations between item attributes and their psychometric properties, and (3) to determine the equivalence of items with similar complexities, and clones of the same item templates.

@@

To ensure data quality, the data from multiple participants were excluded prior to analysis (see Exclusion Criteria below) leaving the data from N=1501 participants for analysis. In these participants, the majority identified as women (men: N=670; women: N=811; non-binary or other: N=13; rather not say: N=2) and the average age was 28.7 years old (sd = 9.9). The sample was relatively well-educated with the majority having completed a Bachelor's degree (N=507) or Master's degree or higher (N=322). Comparatively fewer participants completed only some college (N=471), only a high school degree (N=199), or preferred not to say (N=2).

@@

After providing consent, participants completed 16 items from the MaRs-IB. The design of the MaRs-IB items have been described previously \citep{chierchia2019matrix}. Briefly, each MaRs-IB item consists of a 3 x 3 matrix. Eight of the nine resulting cells contain abstract shapes, while one cell on the bottom right-hand side of the matrix is empty. Participants are instructed to ``complete the matrix'' by identifying the missing shape from among four possible alternatives.

The presentation of each item was preceded by a fixation cross, which lasted for 1200 ms. Upon presentation of the item, participants were given up to 30 s to solve the puzzle. After 25 s elapsed, a clock appeared to count down the remaining 5 s. A trial ended when participants responded, or after 30 s had elapsed without response. Before the trials began, participants reviewed instructions and were made to correctly complete three practice items. Participants were instructed to respond carefully, but to guess if they could not solve the puzzle before the timer ran out. The format of the experiment and instructions were adapted from code publicly released as part of the original study (\url{https://app.gorilla.sc/openmaterials/36164}).

In order to ensure sufficient sampling of every item (and their clones) in the MaRs-IB, participants were administered a pseudorandomly-selected set of 16 out of 64 total items.\footnote{We did not collect new data for the 16 easiest items in the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) because performance on these items was at or near ceiling.} Item sets were constructed as follows: We subdivided the 64 items into 16 sets of four based on their dimensionality (a measure of item complexity defined in \cite{chierchia2019matrix}). Participants were randomly assigned one item from each of the 16 sets. As a consequence, all participants received test forms of roughly equal difficulty. Importantly we also counterbalanced the assignment of item clones across participants, such that we had an approximately uniform number of responses available for each item by shape set (1, 2, or 3) and distractor type (minimal difference, MD, or or paired difference, PD; see \cite{chierchia2019matrix} for details).

@@

The data from multiple participants who completed the experiment were excluded prior to analysis for one or more of the following reasons: rapid guessing (responding in less than 3 s) on four or more trials (N=70); failing to respond on four or more trials (N=8); or minimizing the browser window to a dimension smaller than the minimal requirements (N=7). In total, 83 of 1584 (5.2\%) participants were excluded leaving the data from N=1501 participants available for analysis. Across these participants, each item (64 total) was administered approximately 375 times (sd = 12) and each item clone (384 total) was administered approximately 62 times (sd = 7).

@@

We employed item response models to characterize the psychometric properties of each item. Specifically, we used additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. In AMIS models, item parameters are defined according to a hierarchical structure in which clones descend from item templates. At the core of of all of the AMIS models used here is the three parameter logistic (3PL) item response model, where the probability of a correct response ($y_{ijk} = 1$) for person $i$ on item clone $k$ belonging to item template $j$ is:

@@

\noindent where $\theta_i$ is the latent ability for person $i$, and $\beta_{jk}$, $\alpha_{jk}$, and $\gamma_{jk}$ are the difficulty, discrimination, and guessing parameters for item clone $k$ of item family $j$. Because the estimates of guessing parameters are often unreliable in the absence of very large amounts of data \citep{han2012fixing}, we fixed the guessing parameter for every item clone to the nominal guessing rate ($\gamma_{jk} = 0.25$).

@@

\noindent such that each item difficulty parameter was modeled as an intercept ($\mu_\beta$), reflecting the average difficulty across all items, and four additional components:

@@

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of the template (level 1) attributes on item difficulty, where $Q_{jn}$ is the value of attribute $n$ for item template $j$, and $\delta_{\beta n}$ is the effect of attribute $n$.

\item $\epsilon_{\beta j}$: the template (level 1) residual, or residual variability in item difficulty of the item template unexplained by its attributes.

\item $\sum_{m=1}^M R_{km} \delta_{\beta m}$: the effect of the clone (level 2) attributes on item difficulty, where $R_{km}$ is the value of attribute $m$ for item clone $k$, and $\delta_{\beta m}$ is the effect of attribute $m$.

\item $\epsilon_{\beta k}$: the clone (level 2) residual, or residual variability in item difficulty of the item clone unexplained by its attributes.

@@

\noindent In turn, item discrimination parameters were expressed as the sum of an equivalent set of components:

@@

In our analyses, we considered two template level (level 1) attributes: feature number, or the number of elements in each cell of an item; and rule number, or the number of changing relationships across the cells of an item (e.g. color change, shape change, position change). We predicted that both attributes would be positively associated with item difficulty; that is, all else equal, items with a great number of features or rules would be more difficult to solve. In contrast, we had no hypotheses about the relationship between these attributes and item discrimination.

We also considered two clone level (level 2) attributes: distractor type, or the use of MD-type versus PD-type response alternatives; and mean response time, or the average time participants spent deliberating on that item clone. We note that, in contrast to all other modeled attributes, average response time is not an intrinsic property of an item. We opted to include it in our models, however, because including response time as a covariate has previously been shown to improve parameter estimation \citep{bertling2018using}. To prevent possible collinearity between the attributes, mean response time was orthogonalized with respect to the three other item attributes; therefore, any effects of response time reflects residual structure after accounting for all other item attributes. Based on \cite{chierchia2019matrix}, we predicted there would be no association between distractor type and item difficulty, and a positive association between mean response time and item difficulty. As with the template-level attributes, we had no hypotheses about the relationship between the clone-level attributes and item discrimination.

The AMIS model framework provides a natural means of accomplishing the three goals of the calibration the study. By fitting an AMIS model, we estimate the item parameters and quantify the psychometric properties for every item, and their clones, in the MaRs-IB (Aim 1). By modeling the effects of item attributes, we can evaluate the success of the MaRs-IB design by investigating whether item attributes influence item functioning as expected (e.g. a positive association between rule number and item difficulty; Aim 2). Finally, by estimating the magnitude of residual error in item parameters from unmodeled sources of variance (e.g. perceptual features), we can test whether the MaRs-IB possesses certain desirable structural properties (e.g. psychometric equivalence and exchangeability among item clones; Aim 3).

We also estimated person-level abilities following an explanatory item response modeling approach \citep{wilson2008explanatory}:

@@

We investigated person-level abilities as a function of four covariates: age, gender, average response time, and $\Delta$ response time. This last term reflects the degree of change in a participant's deliberation time as a function of item difficulty (defined as one minus the proportion of correct responses for that item). All covariates were standard-scored prior to analysis with the exception of gender, which was binary coded (male = -0.5, female = 0.5).

@@

During estimation, the item discrimination parameters were restricted to be in the range $\alpha_{jk} \in [0, 5]$. To ensure the identifiability of all models, person abilities were constrained to have a mean of zero and a variance of 1.0. In other words, the residual variance in ability was specified as $V(\epsilon) = 1 - \sum \rho_p^2$.

@@

The goodness of fit of the models was compared using Bayesian leave-one out cross-validation \citep{vehtari2017practical}, which has been found to perform better than more traditional information criteria for comparing item response models \citep{luo2017performances}. We computed the conditional leave-one-cluster out (LOCO) cross-validation \citep{merkle2019bayesian}, which measures a model's ability to generalize to held out items (rather than held out responses or held out participants) --- i.e., for inferring generalization to additional items constructed using the same features rather than to additional participants sampled from the same population.

@@

The fit of the best-fitting model to the data was evaluated using posterior predictive model checking \citep{gelman1996posterior, levy2017bayesian}. A sample of predicted responses was generated for each sample of item parameters and the posterior predictive $p$ (PPP) value was computed based on two discrepancy statistics:  the Orlando and Thissen index ($S-\chi^2$) \citep{toribio2011discrepancy} and the standardized generalized dimensionality discrepancy measure (SGDDM) \citep{levy2015standardized}.

The $S-\chi^2$ discrepancy statistic is a measure of item fit based on the comparison of observed and expected proportions of examinees who answer an item correctly at each raw-score group. A global measure of fit at the test level is obtained by summing the discrepancy values over the items. SGDDM is a measure of model dimensionality. More specifically, SGDDM may be interpreted as the mean absolute conditional correlation between all non-redundant pairs of items, which allows for the identification of inter-item correlations beyond those explained by the model. This latter measure is important as it tests for the presence of non-negligible response dependence across items, which may otherwise lower the efficiency of future test forms due to the inclusion of nonindependent or  redundant items.

@@

On average, participants completed 9.6 of 16 items correctly (sd = 3.1, IQR = 8.0 -- 12.0) (Figure \ref{fig:fig01}a). Timing out occurred on only 1.8\% of trials. A total of 43 participants (2.9\%) performed below chance (i.e. fewer than four items correct), while only 11 participants (0.7\%) solved all 16 of their items. Thus, over 95\% of observed scores were in a reasonable range. These results corroborate those reported by \cite{chierchia2019matrix} in that there were no obvious ceiling or floor effects in performance, and that the majority of participants were sufficiently motivated to participate in spite of the low-stakes testing environment.

Across all 384 item clones, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74; Figure \ref{fig:fig01}b). A total of 12 items (3.1\%) exhibited performance levels beneath chance, and only 22 items (5.7\%) had performance levels near ceiling ($\geq$ 90\%). Across items, the proportion of correct responses was negatively correlated with rule number ($\rho$ = -0.482, p < 0.001). These results further corroborate those reported by \cite{chierchia2019matrix} in that the MaRs-IB exhibited good functioning with performance levels for only a few item clones at ceiling or floor.

We investigated the relationship between accuracy and response time using a mixed effects (random intercepts) linear regression model. Participants' log-transformed response times were predicted as a function of that trial accuracy, rest score (participants' observed scores on all other items), and item difficulty (one minus the proportion correct for that item). The mixed-effects model was estimated using the \textit{statsmodels} python package (v0.12.2) \citep{seabold2010statsmodels}).

The results are summarized in Table \ref{table:1}. On average, participants spent 15.9 s (sd = 7.2 s) on each item. There were significant main effects of accuracy ($\beta = 0.067$, $t = 9.84$, $p < 0.001$), rest score ($\beta = 0.125$, $t = 16.95$, $p < 0.001$), and item difficulty ($\beta = 0.137, t = 44.33, p < 0.001$). That is, response times were slower for correct responses, better-performing participants, and more difficult items.

There were also significant accuracy by difficulty and  accuracy by rest score interactions. Correct responses were even slower for more difficult items, but faster for better performing participants. There was also a significant difficulty by score interaction, such that better performing participants exhibited slower responses for more difficult items. This pattern of response time results is visible in Figure \ref{fig:fig01}c. The three-way interaction term was not significant.

@@

\item Model 1: Item difficulty is a function only of item template attributes (i.e. feature number, rule number). In this model, item with identical attributes are psycho- metrically equivalent, as are all the clones belonging to an item.

\item Model 2: Item difficulty is a function of item template attributes and residuals. In this model, item templates with identical attributes may not be equally difficult; however, all clones belonging to an item are still psychometrically equivalent.
